From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-04-17 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Updates to fundamentality and density section 5
Added in a footnote on aboutness to address Gordon's worry.

Reordered some material in section 5 to be more cogent.

Here, we will argue that there are indeed physical grounds for privileging Hamiltonian mechanics over Lagrangian mechanics, their categorical equivalence notwithstanding.\footnote{Barrett shows that in the hyper-regular domain, Lagrangian and Hamiltonian mechanics are categorically equivalent provided that we define categories of models over the tangent and cotangent bundle formulations, respectively \parencites*[1181-82]{Barrett2}. However, he also shows that if we consider a more general formulation of Hamiltonian mechanics on a symplectic manifold (which does not require a cotangent bundle), then the two formulations are \textit{not} categorically equivalent \parencites*[1182-83]{Barrett2}. Our construction of Hamiltonian mechanics in Section~\ref{derivation} matches this more general symplectic manifold formulation.} Whereas North's argument relies on a problematic criterion for comparing structure (based on the size of symmetry groups) \parencites[]{Swanson}, our argument avoids this pitfall. We show that a system needs to satisfy strictly fewer physical conditions in order to be Hamiltonian. To be Lagrangian, systems must satisfy an additional physical condition, making Lagrangian mechanics a proper subset of Hamiltonian mechanics. Hamiltonian mechanics thereby captures a wider range of systems than Lagrangian mechanics, and this difference has interpretive consequences even within the hyper-regular domain. The construction at the heart of our argument provides an interpretation of classical mechanical systems in terms of their invariant phase space density. We show that the invariant density sits more naturally within the Hamiltonian framework, providing another reason to privilege it over Lagrangian mechanics.
% Our formalism also provides a novel interpretation of the principle of stationary action within Hamiltonian mechanics.

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must clarify what we mean by ``fundamental." Of the various notions of fundamentality, we will mean something entirely non-mysterious: a formalism is more fundamental provided that it is logically more general. In this case, the Lagrangian framework is less general because it requires that an additional condition hold, compared to the Hamiltonian framework. Hence, the Lagrangian framework is less fundamental because it is a proper subset of the Hamiltonian framework.

@@

% % done; slip in a citation to Dorr/Hawthorne 2014, to note the independent metaphysical challenges that the notion of naturalness faces, and likewise for a metaphysical notion of fundamentality.

@@

As indicated at the start of Section~\ref{privileging}, we use ``fundamentality" in a non-metaphysical fashion. On our usage, differences in fundamentality amount to differences in generality.\footnote{We also embrace other non-metaphysical uses of ``fundamental," tracking differences in epistemic or conceptual primacy. For instance, we can think of the canonical position and momentum observables as ``epistemically fundamental" because they suffice to completely specify a classical state, fixing its values for all other observables. \textcites[31, 200]{Ruetsche} refers to this as ``fundamental in the physicist's sense." Section~\ref{fundamental} discusses further senses in which the Hamiltonian formalism is more fundamental than Lagrangian mechanics.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of ``structure." Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of structure, our definition has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which empiricists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

Advantageously, our criterion for comparing Hamiltonian and Lagrangian mechanics avoids the criticisms that Swanson and Halvorson develop against North. Hamiltonian mechanics has less structure than Lagrangian mechanics simply in the sense of logical strength: a system must satisfy strictly fewer physical conditions in order to be Hamiltonian. Whereas to be Lagrangian, the system must also satisfy the assumption of motion equivalence. The assumption of motion equivalence places an additional constraint on the Hamiltonian, a constraint that need not be satisfied in general.

@@

@@

Section~\ref{dependence} shows why point-wise comparison does not hold in general: the density is a coordinate-dependent quantity, and thus it is not generically coordinate-invariant. Consequently, within the Lagrangian formalism, point-wise comparison holds only when we have made a special choice of coordinates. This gives a further reason to privilege Hamiltonian mechanics. Section~\ref{fundamental} argues for the fundamental physical significance of the phase space density. Since the density is coordinate-dependent, this shows that coordinate-independence is not a necessary condition for fundamental physical significance. Our position thereby undercuts metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms.\footnote{For arguments in favor of representing physical systems and explanations in exclusively intrinsic terms, see \textcites[]{Field,North,Chen2018}.} Section~\ref{gauge_momentum} shows why the conjugate momentum $p$ is itself a gauge-dependent quantity. This bolsters our claim that the density is more fundamental than the choice of state variable pairs. It also shows that---\textit{pace} conventional wisdom---position and conjugate momentum are \textit{not} on equal physical footing within Hamiltonian mechanics. The syntactic symmetry between position and momentum in the mathematical formalism does not entail that these quantities are equally physically significant. 
%\footnote{This illustrates how one can be mislead by the syntactical form of the mathematics to draw an interpretive conclusion that is not borne out by the physics.}

@@

The Hamiltonian formalism fixes this issue by assigning areas to regions in a way that is independent of the choice of unit-variables. To do so, it forces us to restrict ourselves to canonical coordinates, which explains why momentum has to change covariantly. The result is that we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in (in the sense of space-time coordinates). In contrast, in the Lagrangian framework, there are at best special coordinate systems that allow for point-wise comparison of the density.
% % Try to precisely gloss/specify the condition for a coordinate system to be special in this way, such that it supports point--wise comparison. (Seems inappropriate to mention the Jacobian, since this matters for transforming from one coordinate system to another.)

@@

* Units are physically significant. In fact, the ability to express the densities in units is what gives us the symplectic structure. It's a physical requirement that we must be able to express the density in some unit.
* Invariant densities: physical significance connected to coordinate invariance. [incorporated]
* Density comparison is independent of coordinates. We can compare densities point-wise without worry. [incorporated]

@@

% Goal of the following paragraph is to explain the various senses in which the invariant phase space density is fundamental

Section~\ref{fundamentality} characterized fundamentality in terms of generality. We argued that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in part because it is more general. By considering how Hamiltonian mechanics supports point-wise density comparisons, we have encountered two further notions of fundamentality. The first is epistemic: by representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all (time-independent) canonical coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See Section~\ref{gauge} for discussion.} Not only is this formal property methodologically advantageous, but it also bespeaks a further physical significance. This brings us to a second, physical notion of fundamentality: the phase space density $\rho$ is physically fundamental because it tracks how much of the system is in each possible state. It quantifies the amount of stuff in each phase space state, in virtue of representing the distribution.

Section~\ref{infinitesimal} demonstrated a third reason for viewing the density as fundamental: by defining the density $\rho_{c, \xi}$, we recover the right mathematical structure, namely a symplectic manifold. If we did not define the density, then we would need another account of why this mathematical structure arises for classical systems. Collectively, these considerations support interpreting the invariant phase space density as not only physically significant but also fundamental. 
%see 11:50am of 12-21 meeting video for clarification of this point. There are probably finer details for me to incorporate.

Yet, as Section~\ref{dependence} explained, the density is always coordinate-dependent, even when coordinate-invariant. In other words, there is no coordinate-independent (or coordinate-free) definition of the invariant phase space density. Affording the density fundamental significance therefore places us at odds with those who take coordinate-independence as a necessary condition for fundamentality.

>>>>

Here, we will argue that there are indeed physical grounds for privileging Hamiltonian mechanics over Lagrangian mechanics, their categorical equivalence notwithstanding.\footnote{Barrett shows that in the hyper-regular domain, Lagrangian and Hamiltonian mechanics are categorically equivalent provided that we define categories of models over the tangent and cotangent bundle formulations, respectively \parencites*[1181-82]{Barrett2}. However, he also shows that if we consider a more general formulation of Hamiltonian mechanics on a symplectic manifold (which does not require a cotangent bundle), then the two formulations are \textit{not} categorically equivalent \parencites*[1182-83]{Barrett2}. Our construction of Hamiltonian mechanics in Section~\ref{derivation} matches this more general symplectic manifold formulation.} Whereas North's argument relies on a problematic criterion for comparing structure based on the size of symmetry groups \parencites[]{Swanson}, our argument avoids this pitfall. We show that a classical system needs to satisfy strictly fewer physical conditions in order to be Hamiltonian. To be Lagrangian, systems must satisfy an additional physical condition, making Lagrangian mechanics a proper subset of Hamiltonian mechanics. Hamiltonian mechanics thereby captures a wider range of systems than Lagrangian mechanics, and this difference has interpretive consequences even within the hyper-regular domain. The construction at the heart of our argument provides an interpretation of classical mechanical systems in terms of their invariant phase space density. We show that the invariant density fits more naturally within the Hamiltonian framework, providing another reason to privilege it over Lagrangian mechanics.
% Our formalism also provides a novel interpretation of the principle of stationary action within Hamiltonian mechanics.

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must clarify what we mean by ``fundamental." Of the various notions of fundamentality, we will mean something entirely non-mysterious: a formalism is more fundamental provided that it is logically more general. In this case, the Lagrangian framework is less general because it requires that an additional condition hold, compared to the Hamiltonian framework. Hence, the Lagrangian framework is less fundamental because it is a proper subset of the Hamiltonian framework.\footnote{One might worry that our criterion for fundamentality is subject to elementary counterexamples. For instance, can't we make any theory more general by adjoining arbitrary disjunctions? Importantly, we require that the theory remains wholly \textit{about} classical mechanics. Hamiltonian mechanics is more general than Lagrangian mechanics while still being about classical systems, where ``aboutness'' can be understood using Yablo's framework \parencites*[]{Yablo}. Having an account of \textit{aboutness} or \textit{subject matter} is a general philosophical problem that any account of fundamentality or essence must address. Hence, it is not a special problem facing our criterion.}
% % What we really want is a brief formalization of ``proper subset," e.g. using partitions of possible worlds semantics. See if I can find Lewis discussion of this (maybe mentioned by Plebani)
% % There is perhaps a further difficulty about the identity criteria for counting conditions. So we really just need it to be the case that all of the Lagrangian systems are a subject matter subset of the Hamiltonian systems. E.g. in the sense that the 1800s is a subset of all of history. need a mereological relationship. 
% Perhaps could say: this notion of a proper subset can be spelled out using Lewis on subject matter. Butterfield/gomes paper. But I could not find this discussion in their paper (might have a false memory)
% % we could formalize this notion of ``proper subset" using what Lewis does regarding the mereology of partitions of possible worlds. This gives an account of subject matters, similar to Yablo's.

@@

@@

As indicated at the start of Section~\ref{privileging}, we use ``fundamentality" in a non-metaphysical fashion. Differences in fundamentality are primarily differences in generality: more fundamental frameworks require fewer conditions to hold.\footnote{We also embrace related non-metaphysical uses of ``fundamental," discussed in Section~\ref{fundamental}.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of ``structure." Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of structure, our definition has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which empiricists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

Advantageously, our criterion for comparing Hamiltonian and Lagrangian mechanics avoids the criticisms that Swanson and Halvorson develop against North. Hamiltonian mechanics has less structure than Lagrangian mechanics simply in the sense of logical strength: a system must satisfy strictly fewer physical conditions in order to be Hamiltonian. Whereas to be Lagrangian, the system must also satisfy the assumption of motion equivalence. The assumption of motion equivalence places an additional constraint on the Hamiltonian, a constraint that need not be satisfied in general. Section~\ref{density} provides a further argument for viewing Hamiltonian mechanics as more fundamental. We argue that (i) invariant phase space densities are a fundamental physical quantity and (ii) Hamiltonian mechanics provides the proper setting for representing these densities.

% % It is also a problem facing the metaphysical accounts of fundamentality, since the notion of ``essence" is also hyperintensional: e.g. Socrates is essentially human vs. Socrates is essentially (human and 2+2 = 4) [example from Stanford page]. So we believe that we are in no worse shape than our more metaphysical opponents.

@@

The phase space density $\rho$ is physically fundamental because it tracks how much of the system is in each possible state. It quantifies the amount of stuff in each phase space state, in virtue of representing the distribution. Typically, the canonical position and momentum observables play this role, since they suffice to completely specify a classical state; they fix the values of all other observables. \textcites[31, 200]{Ruetsche} refers to this as ``fundamental in the physicist's sense." However, we will show in Section~\ref{gauge_momentum} that the conjugate momentum is gauge-redundant. For this reason, the invariant density provides a more fundamental specification of the state than the state variables themselves.\footnote{Section~\ref{infinitesimal} provides another reason for viewing the density as fundamental: by defining the density $\rho_{c, \xi}$, we recover the right mathematical structure, namely a symplectic manifold. If we did not define the density, then we would need another account of why this mathematical structure arises for classical systems.}

@@

Section~\ref{dependence} shows why point-wise comparison does not hold in general: the density is a coordinate-dependent quantity, and thus it is not generically coordinate-invariant. Within the Lagrangian formalism, point-wise comparison holds only when we have made a special choice of coordinates. This gives a further reason to privilege Hamiltonian mechanics. Section~\ref{fundamental} argues that since the density is coordinate-dependent, this shows that coordinate-independence is not a necessary condition for fundamental physical significance. Our position thereby undercuts metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms.\footnote{For arguments in favor of representing physical systems and explanations in exclusively intrinsic terms, see \textcites[]{Field,North,Chen2018}.} Section~\ref{gauge_momentum} shows why the conjugate momentum $p$ is itself a gauge-dependent quantity. This bolsters our claim that the density is more fundamental than the choice of state variable pairs. It also shows that---\textit{pace} conventional wisdom---position and conjugate momentum are \textit{not} on equal physical footing within Hamiltonian mechanics. The syntactic symmetry between position and momentum in the mathematical formalism does not entail that these quantities are equally physically significant. 
%\footnote{This illustrates how one can be mislead by the syntactical form of the mathematics to draw an interpretive conclusion that is not borne out by the physics.}

% % perhaps put at end of section 5.1: We have just shown that Hamiltonian mechanics provides the proper setting for representing the density. By representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in.

@@

The Hamiltonian formalism fixes this issue by assigning areas to regions in a way that is independent of the choice of unit-variables. To do so, it forces us to restrict ourselves to canonical coordinates, which explains why momentum has to change covariantly. The result is that we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in (in the sense of space-time coordinates). Any and all (time-independent) canonical coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See Section~\ref{gauge} for discussion.} In contrast, in the Lagrangian framework, there are at best special coordinate systems that allow for point-wise comparison of the density.
% % Try to precisely gloss/specify the condition for a coordinate system to be special in this way, such that it supports point--wise comparison. (Seems inappropriate to mention the Jacobian, since this matters for transforming from one coordinate system to another.)

@@

* Units are physically significant. In fact, the ability to express the densities in units is what gives us the symplectic structure. It's a physical requirement that we must be able to express the density in some unit. [Sort of transcendental/regulative assumption]
* Invariant densities: physical significance connected to coordinate invariance. [incorporated]
* Density comparison is independent of coordinates. We can compare densities point-wise without worry. [incorporated]

@@

As argued at the start of Section~\ref{density}, the invariant density is physically fundamental because it specifies the state of the classical system. Yet, as Section~\ref{dependence} explained, the density is always coordinate-dependent, even when coordinate-invariant. In other words, there is no coordinate-independent (or coordinate-free) definition of the invariant phase space density. Affording the density fundamental significance therefore places us at odds with those who take coordinate-independence as a necessary condition for fundamentality.

%We have just shown that Hamiltonian mechanics provides the proper setting for representing the density. By representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all (time-independent) canonical coordinate systems will support the formal property of point-wise comparisons.\footnote{Time-dependent coordinate transformations require working in the extended phase space. See Section~\ref{gauge} for discussion.} 
%Collectively, these considerations support interpreting the invariant phase space density as not only physically significant but also fundamental. 
%see 11:50am of 12-21 meeting video for clarification of this point. There are probably finer details for me to incorporate.
